---
title: "Introduction to Bayesian Statistics with R"
subtitle: "3: Monte Carlo methods"
author: "Jack Kuipers"
date: "5 May 2025"
runtime: shiny
output:
  ioslides_presentation:
    css: ibswr.css
    incremental: true
    widescreen: true
    smaller: true
    fig_width: 6
    fig_height: 4.5
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(ggplot2)
```


## Dart throwing

<div id="hidestuffless" class="centered">
```{r circle square}
sidebarPanel(actionButton("start_CP", "Start"), hr(),
  selectInput("n_darts", label = "Number of throws:",
              choices = c(1, 5, 10, 20, 50, 100, 200, 500), selected = 1), hr(),
  sliderInput("wait_timeCP", label = "Wait time:",
              min = 100, max = 1000, value = 600, step = 100), width = 2)

rvCP <- reactiveValues(i = 0, samp_value_x = c(), samp_value_y = c(), flag = FALSE)

observeEvent(input$start_CP, {
  rvCP$i <- 0
  rvCP$samp_value_x <- runif(1, -1, 1)
  rvCP$samp_value_y <- runif(1, -1, 1)
  rvCP$flag <- TRUE
  observe({
    if (isolate(rvCP$flag) == TRUE) {
      isolate({
        rvCP$i <- rvCP$i + 1
        rvCP$samp_value_x <- runif(1, -1, 1)
        rvCP$samp_value_y <- runif(1, -1, 1)      
      })
    }
    if (isolate(rvCP$i) < isolate(as.numeric(input$n_darts))) {
      invalidateLater(input$wait_timeCP, session)
    } else {
      rvCP$flag <- FALSE
    }
  })
})

renderPlot({
  local_samp_value_x <- rvCP$samp_value_x
  local_samp_value_y <- rvCP$samp_value_y
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagFALSE.R")
  source("./ibswrfiles/circle.R", local = TRUE)
}, width = 450, height = 450)
```
</div>

<span class="question">Question: what's the probability your dart hits the dartboard?<br>(given it is uniformly thrown within the square)</span>

## Dart throwing {.build}

<div>
<div id="hidestuffless" class="centered">
```{r circle square many}
sidebarPanel(actionButton("sample_darts", "Sample"), hr(),
  selectInput("n_ds", label = "Number of throws:",
              choices = c(10, 100, 1000, 10000, 100000), selected = 10), width = 2)

renderPlot({
  tempy <- input$sample_darts
  local_samp_value_x <- runif(input$n_ds, -1, 1)
  local_samp_value_y <- runif(input$n_ds, -1, 1)
  hit_flag <- input$show_frac
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagTRUE.R")
  source("./ibswrfiles/circle.R", local = TRUE)
}, width = 450, height = 450)
```
</div>

<span class="answer">Answer: $\frac{\pi}{4} \approx$ `r round(pi/4, 6)` from geometry</span>
</div>

<div class="columns-2">
Or we can estimate by throwing lots of darts!

<p class="forceBreak">&nbsp;</p>

```{r fraction checkbox}
checkboxInput("show_frac", "Show fraction of hits?", value = FALSE)
```

</div>

## Normal samples

<div id="hidestuffless" class="centered">
```{r normal middle}
sidebarPanel(actionButton("start_ns", "Start"), hr(),
  selectInput("n_ns", label = "Number of samples:",
              choices = c(1, 5, 10, 20, 50, 100, 200, 500), selected = 1), hr(),
  sliderInput("wait_time_ns", label = "Wait time:",
              min = 100, max = 1000, value = 600, step = 100), width = 2)

rv_ns <- reactiveValues(i = 0, samp_value = c(), totalCounts = rep(0, 31), flag = FALSE)

observeEvent(input$start_ns, {
  rv_ns$i <- 0
  rv_ns$samp_value <- rnorm(1)
  rv_ns$totalCounts = rep(0, 31)
  rv_ns$flag <- TRUE
  observe({
    if (isolate(rv_ns$flag) == TRUE) {
      isolate({
        rv_ns$i <- rv_ns$i + 1
        rv_ns$samp_value <- rnorm(1)
        hitIndex <- floor(rv_ns$samp_value*5) + 16
        if(hitIndex > 0 && hitIndex < 31){
          rv_ns$totalCounts[hitIndex] <- rv_ns$totalCounts[hitIndex] + 1
        } else { # put in rubbish bin
          rv_ns$totalCounts[31] <- rv_ns$totalCounts[31] + 1
        }
      })
    }
    if (isolate(rv_ns$i) < isolate(as.numeric(input$n_ns))) {
      invalidateLater(input$wait_time_ns, session)
    } else {
      rv_ns$flag <- FALSE
    }
  })
})
  
renderPlot({
  local_samp_value <- rv_ns$samp_value
  local_counts <- rv_ns$totalCounts
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagTRUE.R")
  source("./ibswrfiles/normalarea.R", local = TRUE)
}, width = 600, height = 450)
```
</div>

<span class="question">Question: what's the probability a standard normal sample is between -1 and 1?</span>


```{r cumulative checkbox}
checkboxInput("show_cum", "Show cumulative hits?", value = FALSE)
```


## Normal samples

<div>
<div id="hidestuffless" class="centered">
```{r normal middle many}
sidebarPanel(actionButton("sample_ns", "Sample"), hr(),
  selectInput("n_norms", label = "Number of samples:",
              choices = c(10, 100, 1000, 10000, 100000, 1e6), selected = 10), width = 2)

renderPlot({
  tempy <- input$sample_ns
  local_samp_value <- rnorm(input$n_norms)
  local_samp_value[which(local_samp_value < -3)] <- 3.1
  local_samp_value[which(local_samp_value > 3)] <- 3.1
  local_counts <- tabulate(floor(local_samp_value*5)+16, nbins = 31)
  area_flag <- input$show_area
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagFT.R")
  source("./ibswrfiles/normalarea.R", local = TRUE)
}, width = 600, height = 450)
```
</div>

<span class="answer">Answer: `r round(pnorm(1)-pnorm(-1), 6)` from calculus (next slide)</span>
</div>

<div class="columns-2">
Or we can estimate from lots of random samples!

<p class="forceBreak">&nbsp;</p>

```{r area checkbox}
checkboxInput("show_area", "Show fraction inside?", value = FALSE)
```

</div>

## Integration

<div class="columns-2">
We are actually integrating

$$\int_{-1}^{1} f(x) \mathrm{d}(x)$$

with $f(x)$ the probability density of a standard normal

$$f(x) = \frac{1}{\sqrt{2\pi}}\mathrm{e}^{-\frac{x^2}{2}}$$

Analytical result is $\Phi(1) - \Phi(-1)$

>- $\Phi$ is CDF of a normal (an error function)  

```{r, echo = TRUE}
pnorm(1) - pnorm(-1)
```

<p class="forceBreak">&nbsp;</p>

<div class="centered">
```{r normal num one}
renderPlot({
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagFALSE.R")
  source("./ibswrfiles/normalareanum.R", local = TRUE)
}, width = 400, height = 250)
```
</div>

</div>

## Numerical integration

<div class="columns-2">
Let's approximate the integral

$$\int_{-1}^{1} f(x) \mathrm{d}(x)$$

with lots of rectangular bins:

$$ \approx \sum_{i=1}^{B} f(x_i) \delta x$$

>- $B$ bins
>- $x_i$ are the midpoints of the bins
>- $\delta x$ is the width of the bins

Numerical result: (*cf* `r round(pnorm(1)-pnorm(-1), 6)`)

```{r}
renderUI(round(rv_int$int_result, 6))
```

<p class="forceBreak">&nbsp;</p>

<div class="centered">
```{r normal num two}
renderPlot({
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagTRUE.R")
  n_bins <- as.numeric(input$n_norm_bins)
  source("./ibswrfiles/normalareanum.R", local = TRUE)
}, width = 400, height = 250)
```

```{r}
inputPanel(selectInput("n_norm_bins", label = "Number of bins:",
            choices = c(1, 5, 10, 20, 50, 100), selected = 10))

rv_int <- reactiveValues(int_result = 0)

observeEvent(input$n_norm_bins, {
  n_bins <- as.numeric(input$n_norm_bins)
  bs <- seq(-1+1/n_bins, 1-1/n_bins, length.out=n_bins)
  rv_int$int_result <- 2*sum(dnorm(bs))/(n_bins)
  }
)
```
</div>

</div>



## Sampling approximation

<div class="columns-2">
Or we can approximate the integral

$$\int_{-1}^{1} f(x) \mathrm{d}(x)$$

with lots of normal samples:

$$ \approx \frac{1}{M}\sum_{i=1}^{M} I(-1< x_i < 1)$$

>- $M$ samples 
>- $x_i$ are the samples $\sim f(x)$
>- $I$ is the indicator function

Numerical result: (*cf* `r round(pnorm(1)-pnorm(-1), 6)`)

```{r}
renderUI(round(rv_int_s$int_result, 6))
```

<p class="forceBreak">&nbsp;</p>

<div class="centered">
```{r normal num three}
renderPlot({
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagsTRUE.R")
  local_tables <- rv_int_s$tables
  n_bins <- length(local_tables)
  source("./ibswrfiles/normalareanum.R", local = TRUE)
}, width = 400, height = 250)
```

```{r}
inputPanel(selectInput("n_norm_samps", label = "Number of samples:",
            choices = c(1000, 10000, 100000, 1e6), selected = 1000),
actionButton("n_norm_sample", "Resample"))

rv_int_s <- reactiveValues(int_result = 0, tables = rep(0, 10))

observeEvent({
  input$n_norm_samps
  input$n_norm_sample
  }, {
  if (is.null(input$n_norm_samps)) {
    r_samps <- rnorm(1000)
  } else {
    r_samps <- rnorm(as.numeric(input$n_norm_samps))
  }
  n_bins <- 5*(log10(length(r_samps))-2)
  in_middle <- which(r_samps < 1 & r_samps > -1)
  rv_int_s$tables <- n_bins*tabulate(floor((r_samps[in_middle]+1)*n_bins/2)+1, n_bins)/(2*length(r_samps))
  rv_int_s$int_result <- length(in_middle)/length(r_samps)
  }, ignoreNULL = FALSE
)
```
</div>

</div>


## More integration

<div class="columns-2">
Let's now integrate

$$\int_{-1}^{1} x^2 f(x) \mathrm{d}(x)$$
with lots of rectangular bins

$$ \approx \sum_{i=1}^{B} x_i^2 f(x_i) \delta x$$

Numerical result:

```{r}
renderUI(round(rv_int_x2$int_result, 6))
```

<br> 

The analytical result is $\Phi(1) - \Phi(-1) - 2f(1)$

```{r, echo = TRUE}
pnorm(1) - pnorm(-1) - 2*dnorm(1)
```


<p class="forceBreak">&nbsp;</p>

<div class="centered">
```{r x2 normal num one}
renderPlot({
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagTRUE.R")
  n_bins <- as.numeric(input$n_norm_binsx2)
  source("./ibswrfiles/x2normalareanum.R", local = TRUE)
}, width = 400, height = 250)
```

```{r}
inputPanel(selectInput("n_norm_binsx2", label = "Number of bins:",
            choices = c(5, 10, 20, 50, 100, 200), selected = 10))

rv_int_x2 <- reactiveValues(int_result = 0)

observeEvent(input$n_norm_binsx2, {
  n_bins <- as.numeric(input$n_norm_binsx2)
  bs <- seq(-1+1/n_bins, 1-1/n_bins, length.out=n_bins)
  rv_int_x2$int_result <- 2*sum(dnorm(bs)*bs^2)/(n_bins)
  }
)
```
</div>

</div>


## More integration split

<div id ="hidestuffmore" class="columns-2">

<br><br><br>

Let's now integrate

$$\int_{-1}^{1} x^2 f(x) \mathrm{d}(x)$$

with lots of rectangular bins

$$ \approx \sum_{i=1}^{B} x_i^2 f(x_i) \delta x$$

Numerical result: (*cf* `r round(pnorm(1) - pnorm(-1) - 2*dnorm(1), 6)`)

```{r}
renderUI(round(rv_int_x2two$int_result, 6))
```

<br>

Same as multiplying the bins from normal

>- by $x_i^2$ at their midpoints

$$ x_i^2 \cdot f(x_i) \delta x$$


<p class="forceBreak">&nbsp;</p>

<div class="centered">
```{r x2 normal num split x}
renderPlot({
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagTRUE.R")
  n_bins <- as.numeric(input$n_norm_binsx2two)
  source("./ibswrfiles/x2curvenum.R", local = TRUE)
}, width = 400, height = 150)

renderPlot({
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagTRUE.R")
  n_bins <- as.numeric(input$n_norm_binsx2two)
  source("./ibswrfiles/normalareanum.R", local = TRUE)
}, width = 400, height = 250)
```

```{r}
inputPanel(selectInput("n_norm_binsx2two", label = "Number of bins:",
            choices = c(10, 20, 50, 100), selected = 10))

rv_int_x2two <- reactiveValues(int_result = 0)

observeEvent(input$n_norm_binsx2two, {
  n_bins <- as.numeric(input$n_norm_binsx2two)
  bs <- seq(-1+1/n_bins, 1-1/n_bins, length.out=n_bins)
  rv_int_x2two$int_result <- 2*sum(dnorm(bs)*bs^2)/(n_bins)
  }
)
```
</div>

</div>


## More integration sampling

<div id ="hidestuffmore" class="columns-2">

<br><br><br>

For our integral

$$\int_{-1}^{1} x^2 f(x) \mathrm{d}(x)$$

we can replace the normal bins

>- with samples from the normal

$$ \approx \frac{1}{M}\sum_{i=1}^{M} x_i^2 I(-1< x_i < 1)$$

Since the $x_i$ are sampled $\sim f(x)$

>- we just need to average $x^2$ over the sample

Numerical result: (*cf* `r round(pnorm(1) - pnorm(-1) - 2*dnorm(1), 6)`)

```{r}
renderUI(round(rv_int_sx2$int_result, 6))
```

<p class="forceBreak">&nbsp;</p>

<div class="centered">
```{r x2 normal num split two}
renderPlot({
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagFALSE.R")
  source("./ibswrfiles/x2curvenum.R", local = TRUE)
}, width = 400, height = 150)

renderPlot({
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagsTRUE.R")
  local_tables <- rv_int_sx2$tables
  n_bins <- length(local_tables)
  source("./ibswrfiles/normalareanum.R", local = TRUE)
}, width = 400, height = 250)
```

```{r}
inputPanel(selectInput("n_norm_sampsx2", label = "Number of samples:",
            choices = c(1000, 10000, 100000, 1e6), selected = 1000),
actionButton("n_norm_samplex2", "Resample"))

rv_int_sx2 <- reactiveValues(int_result = 0, tables = rep(0, 10))

observeEvent({
  input$n_norm_sampsx2
  input$n_norm_samplex2
  }, {
  if (is.null(input$n_norm_sampsx2)) {
    r_samps <- rnorm(1000)
  } else {
    r_samps <- rnorm(as.numeric(input$n_norm_sampsx2))
  }
  n_bins <- 5*(log10(length(r_samps))-2)
  in_middle <- which(r_samps < 1 & r_samps > -1)
  rv_int_sx2$tables <- n_bins*tabulate(floor((r_samps[in_middle]+1)*n_bins/2)+1, n_bins)/(2*length(r_samps))
  rv_int_sx2$int_result <- sum(r_samps[in_middle]^2)/length(r_samps)
  }, ignoreNULL = FALSE
)
```

</div>

</div>


## Monte Carlo approximation {.build}

<div>
In general, these integrals are <span class="def">expectations</span>

$$E[g(x)] = \int g(x) p(x) \mathrm{d} x$$
</div>

<div>
If we can sample proportionally to $p(x)$

>- can use <span class="def">Monte Carlo</span> approximation

$$E[g(x)] \approx \frac{1}{M} \sum_{i=1}^{M} g(x_i) $$

>- for $M$ random samples $x_i$, sampled $\sim p(x)$
</div>

<div>
Accuracy typically scales like $\frac{1}{\sqrt{M}}$ for any dimension

>- number of bins for numerical integration grows exponentially in the dimension
</div>

<div>
Lies behind lots of Bayesian analyses

<span class="question">Question: how do we get samples from arbitrary distributions?</span>
</div>


## MCMC {.build}

Markov Chain Monte Carlo (MCMC) **Metropolis, Rosenbluth**$\times$**2, Teller**$\times$**2 (1953); Hastings  (1970)**

<div>
To sample proportionally to any function $p(x)$:
</div>

<div>
[1.] From current position $x$, propose a new position $x'$

$$x' = x + \delta\, , \qquad \delta \sim q(d)$$

>- with a symmetric distribution $q$(d), *eg* a normal (can modify algorithm for asymmetric too)
</div>

<div>
[2.] Accept $x'$ with probability

$$\min \left(1, \frac{p(x')}{p(x)}\right)$$

>- and move current position to $x'$
>- otherwise stay at $x$
</div>

<div>
[3.] Repeat, that's all!

>- (though it may take many iterations to converge)
</div>


## MCMC example

```{r}
sidebarPanel(actionButton("mcmc_step", "Next step"), width = 2)

target_dens <- function(x) {
  0.6*dnorm(x, -0.6, 0.6/1.6) + 0.4*dnorm(x, 0.4, 0.4/1.6)
}

q_sd <- 0.42

rv_mcmc <- reactiveValues(x = 0, 
                          x_p = 0,
                          mh_rat = 0,
                          acc_flag = 0)

observeEvent(input$mcmc_step, {
  if (rv_mcmc$acc_flag) {
    rv_mcmc$x <- rv_mcmc$x_p
  }
  rv_mcmc$x_p <- rv_mcmc$x + rnorm(1, 0, q_sd)
  rv_mcmc$mh_rat <- min(1, target_dens(rv_mcmc$x_p)/target_dens(rv_mcmc$x))
  rv_mcmc$acc_flag <- 1*(runif(1) < rv_mcmc$mh_rat)
  }, ignoreNULL = FALSE, ignoreInit = TRUE
)

renderPlot({
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagFALSE.R")
  cur_x <- rv_mcmc$x
  x_p <- rv_mcmc$x_p
  acc_flag <- rv_mcmc$acc_flag
  source("./ibswrfiles/mcmc_ex.R", local = TRUE)
}, width = 800, height = 400)
```

```{r}
renderUI(paste0("Current x: ", round(rv_mcmc$x, 6)))
renderUI(paste0("Proposed x: ", round(rv_mcmc$x_p, 6)))
```

```{r}
renderUI(paste0("Acceptance probability: ", round(rv_mcmc$mh_rat, 6)))
renderUI(paste0("Move was ", ifelse(rv_mcmc$acc_flag == 1, "accepted", "rejected"), "."))
```


## MCMC samples

<div id="hidestuffless" class="centered">
```{r mcmc chain}
sidebarPanel(actionButton("start_mcmc", "Start"), hr(),
  selectInput("n_mcmc", label = "Number of samples:",
              choices = c(1, 10, 40, 100, 400, 1000), selected = 1), hr(),
  sliderInput("wait_time_mcmc", label = "Wait time:",
              min = 50, max = 500, value = 500, step = 50), width = 2)

rv_mcmc2 <- reactiveValues(i = 0, x = 0, x_p = 0, mh_rat = 0, acc_flag = 0,
                           totalCounts = rep(0, 31), flag = FALSE)

observeEvent(input$start_mcmc, {
  rv_mcmc2$i <- 0
  rv_mcmc2$x_p <- rv_mcmc2$x + rnorm(1, 0, q_sd)
  rv_mcmc2$mh_rat <- min(1, target_dens(rv_mcmc2$x_p)/target_dens(rv_mcmc2$x))
  rv_mcmc2$acc_flag <- 1*(runif(1) < rv_mcmc2$mh_rat)
  rv_mcmc2$totalCounts = rep(0, 31)
  rv_mcmc2$flag <- TRUE
  observe({
    if (isolate(rv_mcmc2$flag) == TRUE) {
      isolate({
        rv_mcmc2$i <- rv_mcmc2$i + 1
        if (rv_mcmc2$acc_flag) {
          rv_mcmc2$x <- rv_mcmc2$x_p
        }
        rv_mcmc2$x_p <- rv_mcmc2$x + rnorm(1, 0, q_sd)
        rv_mcmc2$mh_rat <- min(1, target_dens(rv_mcmc2$x_p)/target_dens(rv_mcmc2$x))
        rv_mcmc2$acc_flag <- 1*(runif(1) < rv_mcmc2$mh_rat)
        hitIndex <- floor(rv_mcmc2$x*5) + 16
        if(hitIndex > 0 && hitIndex < 31){
          rv_mcmc2$totalCounts[hitIndex] <- rv_mcmc2$totalCounts[hitIndex] + 1
        } else { # put in rubbish bin
          rv_mcmc2$totalCounts[31] <- rv_mcmc2$totalCounts[31] + 1
        }
      })
    }
    if (isolate(rv_mcmc2$i) < isolate(as.numeric(input$n_mcmc))) {
      invalidateLater(input$wait_time_mcmc, session)
    } else {
      rv_mcmc2$flag <- FALSE
    }
  })
})
  
renderPlot({
  cur_x <- rv_mcmc2$x
  x_p <- rv_mcmc2$x_p
  acc_flag <- rv_mcmc2$acc_flag
  local_counts <- rv_mcmc2$totalCounts
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagTRUE.R")
  source("./ibswrfiles/mcmc_ex.R", local = TRUE)
}, width = 600, height = 450)
```
</div>

## HMC {.build}

Hamiltonian Monte Carlo (HMC) **Duane, Kennedy, Pendleton and Roweth (1987); Neal (1995, 2011)**

<div>
To sample proportionally to any function $p(x)$: add a new dimension $\rho$!
</div>

<div>
Make $\rho$ Gaussian, and define the negative log-likelihood as an energy surface
$$H(x,\rho) = -\log\{p(x)\} + \frac{\rho^2}{2}$$
</div>

<div>
[1.] From current position $x$, sample a Gaussian $\rho$ and propose a new pair $(x', \rho')$ by propagating (numerically) under the (energy-preserving) Hamiltonian dynamics: $\frac{\mathrm{d}x}{\mathrm{d}t} = \frac{\partial H}{\partial \rho}, \,   \frac{\mathrm{d}\rho}{\mathrm{d}t} = -\frac{\partial H}{\partial x}$
</div>

<div>
[2.] Accept $x'$ with probability

$$\min \left(1, \exp\{H(x,\rho) - H(x', \rho')\}\right) \approx 1$$

>- and move current position to $x'$
>- otherwise stay at $x$
</div>

<div>
[3.] Repeat, that's all!
</div>


## HMC example

```{r}
sidebarPanel(actionButton("hmc_step", "Next step"), width = 2)

target_dens <- function(x) {
  0.6*dnorm(x, -0.6, 0.6/1.6) + 0.4*dnorm(x, 0.4, 0.4/1.6)
}

U <- function(x) {
  part1 <- 0.6*dnorm(x, -0.6, 0.6/1.6)
  part2 <- 0.4*dnorm(x, 0.4, 0.4/1.6)
  d1 <- -(x+0.6)/(0.6/1.6)^2*part1
  d2 <- -(x-0.4)/(0.4/1.6)^2*part2
  dist <- part1 +  part2
  grad <- d1 + d2
  return(list(U = -log(dist), grad = -grad/dist))
}

xs <- 10*c(-300:300)/1000
ys <- target_dens(xs)
yscale <- max(ys)

rhos <- c(-100:100)/20
z <- outer(-log(ys), rhos^2/2, "+")

q_sd <- 0.42

L <- 10

rv_hmc <- reactiveValues(x = 0, 
                         x_p = 0,
                         p = 0,
                         p_p = 0,
                         xs = rep(0,L+1),
                         ps = rep(0,L+1),
                         mh_rat = 0,
                         acc_flag = 0)

observeEvent(input$hmc_step, {
  if (rv_hmc$acc_flag) {
    rv_hmc$x <- rv_hmc$x_p
  }
  q <- rv_hmc$x
  p <- rnorm(1,0,1)
  rv_hmc$p <- p
  p0 <- p
  epsilon <- 0.1
  qs <- rep(0,L+1)
  ps <- rep(0,L+1)
  qs[1] <- q
  U_q <- U(q)
  p <- p - epsilon / 2 * U_q$grad
  ps[1] <- p
    for (j in 1:L) {
      # Make a full step for the position
      q <- q + epsilon * p 
      # Make a full step for the momentum
      # Make a half step for momentum at the end
      U_q_new <- U(q)
      p <- p - epsilon * U_q_new$grad/(1 + (j==L))
      qs[j+1] <- q
      ps[j+1] <- p
    }
  rv_hmc$xs <- qs
  rv_hmc$ps <- c(p0, ((ps[-(L+1)]+ps[-1])/2)[-L], p)
  rv_hmc$x_p <- q
  rv_hmc$p_p <- p
  rv_hmc$mh_rat <- min(1, exp(U_q$U + p0^2/2 - U_q_new$U - p^2/2))
  rv_hmc$acc_flag <- 1*(runif(1) < rv_hmc$mh_rat)
  }, ignoreNULL = FALSE, ignoreInit = TRUE
)

renderPlot({
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagFALSE.R")
  cur_x <- rv_hmc$x
  x_p <- rv_hmc$x_p
  acc_flag <- rv_hmc$acc_flag
  step_rhos <- rv_hmc$ps
  step_xs <- rv_hmc$xs
  source("./ibswrfiles/hmc_ex.R", local = TRUE)
}, width = 800, height = 400)
```

```{r}
renderUI(paste0("Current x: ", round(rv_hmc$x, 6), ", random \u03c1: ", round(rv_hmc$p, 6)))
renderUI(paste0("Proposed x: ", round(rv_hmc$x_p, 6), ", proposed \u03c1: ", round(rv_hmc$p_p, 6)))
```

```{r}
renderUI(paste0("Acceptance probability: ", round(rv_hmc$mh_rat, 6)))
renderUI(paste0("Move was ", ifelse(rv_hmc$acc_flag == 1, "accepted", "rejected"), "."))
```


## HMC samples

<div id="hidestuffless" class="centered">
```{r hmc chain}
sidebarPanel(actionButton("start_hmc", "Start"), hr(),
  selectInput("n_hmc", label = "Number of samples:",
              choices = c(1, 10, 40, 100, 400, 1000), selected = 1), hr(),
  sliderInput("wait_time_hmc", label = "Wait time:",
              min = 50, max = 500, value = 500, step = 50), width = 2)

rv_hmc2 <- reactiveValues(i = 0, x = 0, x_p = 0, 
                          p = 0, p_p = 0, xs = rep(0,L+1), ps = rep(0,L+1),
                         mh_rat = 0, acc_flag = 0,
                           totalCounts = rep(0, 31), flag = FALSE)

observeEvent(input$start_hmc, {
  rv_hmc2$i <- 0
  q <- rv_hmc2$x
  p <- rnorm(1,0,1)
  rv_hmc2$p <- p
  p0 <- p
  epsilon <- 0.1
  qs <- rep(0,L+1)
  ps <- rep(0,L+1)
  qs[1] <- q
  U_q <- U(q)
  p <- p - epsilon / 2 * U_q$grad
  ps[1] <- p
    for (j in 1:L) {
      # Make a full step for the position
      q <- q + epsilon * p 
      # Make a full step for the momentum
      # Make a half step for momentum at the end
      U_q_new <- U(q)
      p <- p - epsilon * U_q_new$grad/(1 + (j==L))
      qs[j+1] <- q
      ps[j+1] <- p
    }
  rv_hmc2$xs <- qs
  rv_hmc2$ps <- c(p0, ((ps[-(L+1)]+ps[-1])/2)[-L], p)
  rv_hmc2$x_p <- q
  rv_hmc2$p_p <- p
  rv_hmc2$mh_rat <- min(1, exp(U_q$U + p0^2/2 - U_q_new$U - p^2/2))
  rv_hmc2$acc_flag <- 1*(runif(1) < rv_hmc2$mh_rat)
  rv_hmc2$totalCounts = rep(0, 31)
  rv_hmc2$flag <- TRUE
  observe({
    if (isolate(rv_hmc2$flag) == TRUE) {
      isolate({
        rv_hmc2$i <- rv_hmc2$i + 1
        if (rv_hmc2$acc_flag) {
          rv_hmc2$x <- rv_hmc2$x_p
        }
  q <- rv_hmc2$x
  p <- rnorm(1,0,1)
  rv_hmc2$p <- p
  p0 <- p
  epsilon <- 0.1
  qs <- rep(0,L+1)
  ps <- rep(0,L+1)
  qs[1] <- q
  U_q <- U(q)
  p <- p - epsilon / 2 * U_q$grad
  ps[1] <- p
    for (j in 1:L) {
      # Make a full step for the position
      q <- q + epsilon * p 
      # Make a full step for the momentum
      # Make a half step for momentum at the end
      U_q_new <- U(q)
      p <- p - epsilon * U_q_new$grad/(1 + (j==L))
      qs[j+1] <- q
      ps[j+1] <- p
    }
  rv_hmc2$xs <- qs
  rv_hmc2$ps <- c(p0, ((ps[-(L+1)]+ps[-1])/2)[-L], p)
  rv_hmc2$x_p <- q
  rv_hmc2$p_p <- p
  rv_hmc2$mh_rat <- min(1, exp(U_q$U + p0^2/2 - U_q_new$U - p^2/2))
        rv_hmc2$acc_flag <- 1*(runif(1) < rv_hmc2$mh_rat)
        hitIndex <- floor(rv_hmc2$x*5) + 16
        if(hitIndex > 0 && hitIndex < 31){
          rv_hmc2$totalCounts[hitIndex] <- rv_hmc2$totalCounts[hitIndex] + 1
        } else { # put in rubbish bin
          rv_hmc2$totalCounts[31] <- rv_hmc2$totalCounts[31] + 1
        }
      })
    }
    if (isolate(rv_hmc2$i) < isolate(as.numeric(input$n_hmc))) {
      invalidateLater(input$wait_time_hmc, session)
    } else {
      rv_hmc2$flag <- FALSE
    }
  })
})
  
renderPlot({
  source("./ibswrfiles/colordefs.R")
  source("./ibswrfiles/flagTRUE.R")
  cur_x <- rv_hmc2$x
  x_p <- rv_hmc2$x_p
  acc_flag <- rv_hmc2$acc_flag
  step_rhos <- rv_hmc2$ps
  step_xs <- rv_hmc2$xs
  local_counts <- rv_hmc2$totalCounts
  source("./ibswrfiles/hmc_ex.R", local = TRUE)
}, width = 600, height = 450)
```
</div>


## Summary {.build}

<div>
With random samples can obtain <span class="def">Monte Carlo</span> estimates

>- to expectations and integrals
</div>

<div>
Behind the scenes in most <span class="def">Bayesian</span> analyses
</div>

<div>
To obtain random samples estimates

>- can use <span class="def">Markov Chain Monte Carlo</span> (MCMC)
</div>

<div>
Start of chain correlated with initial point

>- often remove initial <span class="def">burn-in</span> iterations
</div>

<div>
Faster, more computationally efficient schemes

>- <span class="def">Hamiltonian Monte Carlo</span> (HMC)
</div>

<div>
$\rightarrow$ Exercises 3
</div>




