---
title: "Introduction to Bayesian Statistics with R"
subtitle: "4: Exercises"
author: "Jack Kuipers"
date: "5 May 2025"
output:
  pdf_document:
    fig_width: 6
    fig_height: 3
    latex_engine: xelatex
    includes:
      in_header: ./logos/header.tex
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, include = TRUE)
library(ggplot2)
```

## Exercise 4.1 - Credible intervals

The likelihood of data $D$ made up of $N$ observations of $x$: $x_i, i = 1,\ldots, N$, under a normal model with mean $\mu$ and sd $\sigma$ is

$$P(D \mid \mu, \sigma) \propto 
\prod_{i = 1}^{N}\frac{1}{\sigma}\mathrm{e}^{-\frac{(x_i-\mu)^2}{2\sigma^2}} =
\frac{1}{\sigma^{N}}\mathrm{e}^{-\frac{\sum_{i=1}^{N} (x_i-\mu)^2}{2\sigma^2}} = 
\frac{1}{\sigma^{N}}\mathrm{e}^{-\frac{1}{2\sigma^2}\left[\sum_{i=1}^{N} x_i^2 -2\mu\sum_{i=1}^{N} x_i + N\mu^2\right]}$$

so that, by precomputing the sum of $x_i^2$ and $x_i$ we can evaluate the likelihood with the following functions

```{r, eval = FALSE}
x2 <- sum(x^2)
x1 <- sum(x)
x0 <- length(x) # number of observations
g_like <- function(mu, sigma, x2, x1, x0) {
  exp(-(x2 - 2*mu*x1 + x0*mu^2)/(2*sigma^2))/(sigma^x0)
}
```

**Note:** for numerical accuracy we often log-transform and work with the log-likelihood.

For now we keep $\sigma=1$ fixed and have a Student-$t$ prior on $\mu$ with scale $t_s$ and $\nu$ degrees of freedom

```{r, eval = FALSE}
prior_t <- function(mu, t_s, nu) {
  dt(mu/t_s, nu)/t_s
}
```

Update the \code{basicMCMC} code from Exercise 3.1 to sample $\mu$ from its posterior distribution, depending on the data $x$, and the prior choices $t_s$ and $\nu$.

For data we will use Gosset's data on barley seed yield from his 1908 paper which introduced the Student-$t$ distribution. The yields (in US cwt per acre) for 11 plots of land for normal seed and seed which has been dried in a kiln are in \code{seed\_data.csv}.

* Read in the data, and extract the relevant quantity as $x$
* Choose prior parameters $t_s$ and $\nu$
* Sample from the posterior distribution of $\mu$
* Visualise this distribution
* Obtain estimates for the 95% credible interval of $\mu$

